---
title: "R Markdown for data exploration homework"
output: github_document
---

## Setting up

The data we'll analyze is contained in the `gapminder` R package. If not installed, run `install.packages("gapminder")` first.

```{r}
library(gapminder)
```


## Exploration of data frame

### Basic info of the data frame
Basic functions like `str`, `names`, `ncol`, `nrow`, `summary` help to get an overview of the data frame.

```{r}
str(gapminder) #explores the structure of the data

```

We have 6 variables: *country*, *continent*, *year*, *lifeExp* (life expentancy), *pop* (population), and *gdpPercap* (GDP per capita). Some are **numerical** and some are **categorical** (factors)

We can also see the variables using the function `names`

```{r}
names(gapminder)
```

`ncol` and `nrow` will give us the number of columns and rows respectively

```{r}
ncol(gapminder)
nrow(gapminder)
```

Perhaps the most useful basic function is `summary`

```{r}
summary(gapminder)
```

### Plots

Plots help visualizing data,  we use `plot` to make basic graphs in R

```{r}
plot(lifeExp ~ year, gapminder)
plot(lifeExp ~ gdpPercap, gapminder)
```

### Extracting columns or variables

To specify a single variable from a data frame, use the dollar sign $. We can extract a single variable and save the data in a new vector (not very useful in this case though).

```{r}
LE <- gapminder$lifeExp # saves the vector of life expectancy
```

Now we know how to extract a variable, we can plot variables more easily.
Lets try a couple more types of plots

### Histograms
Another basic kind of plot are the histograms, for these we use `hist`.
Let’s explore the numeric variable for life expectancy.

```{r}
summary(gapminder$lifeExp)
hist(gapminder$lifeExp)

```

### Bar plots

Using `str` we see the variables *continent* and *country* are categorical (factors). We can try plotting the number of countries in each continent with a bar plot.

```{r}
str(gapminder$continent)
barplot(table(gapminder$continent))
```
 
#### Boxplots

Let's also try a `boxplot` to look at the life expentacy by continent. this shows the median, lower and upper quartiles and the max and min values.

```{r}
boxplot(gapminder$lifeExp ~ gapminder$continent)
```

